Methods for estimating parasite-induced mortality from intensity data and their limitations

### 2015-05-11 docs/parasite_mortatlity_manuscript.tex
Added Figure 3 simulations and Saras commentts

\title{Comparing methods for estimating parasite-induced host mortality from distributional data: implementation and limitations}
\author{Mark Wilber, Sara Weinstein, and Cherie Briggs}

@@

Infectious agents can have major impacts on animal populations through changing
population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies(), bovine TB(), and
rinderpest(),  is typically quantified based on the presence or absence of
disease, and does not account for the number of infectious agents present.
Although sufficient for many bacterial and viral agents that reproduce within a
host, for macroparasites, hosts cannot be simply categorized as infected and
uninfected because pathology is linked to the intensity of infection \citep{AndersonandMay1979}.  Helminths exhibiting this intensity dependent pathology have
significant impacts on human health (), domestic livestock economics (),
wildlife survival ().  While it is generally assumed that some fraction of
wild host populations must succumb to parasitic infections, it is notoriously
difficult to actually quantify parasite-induced host mortality (PIHM) in wild
animal populations \citep{Lafferty2002, McCallum2000a}.

Ideally, parasite-induced host mortality would be
quantified by experimentally infecting and tracking individual hosts in the
wild population; however, for logistical and ethical reasons this method is
rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
often been used to identify the presence PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
relationship between infection intensity and host mortality \citep{Adjei1986} [others?].

\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
observed parasite distribution in the host population to the distribution
predicted in the absence of parasite-induced host mortality. This method
(“Crofton Method”) assumes that, prior to host mortality, parasites are
distributed in the host population following a negative binomial distribution;
however, as intensity dependent pathology removes heavily infected hosts from
the population, the tail of the distribution is truncated. Mortality is assumed
to not occur in hosts with low intensity infections, thus by iteratively
fitting a negative binomial distribution to hosts with lower and lower parasite
loads, and comparing the tail end of this predicted distribution to the
observed parasite data, one could determine both whether PIHM was occurring and
the parasite distribution in the host population prior to parasite induced
mortality. We give a thorough description and implementation of the Crofton
Method in \emph{Supplementary Information (SI)} 1 and discuss the validity of
its assumptions in the \emph{Discussion}.

The Crofton Method may be able to detect the presence of PIHM however,
quantifying the relationship between infection intensity and host survival
probability is more complicated. \cite{Adjei1986} suggested that these values
could be quantified by using the Crofton Method to first estimate the pre-
mortality parasite distribution and then, using those parameters, calculate the
probability of host survival with increasing parasite load. To do this,
\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI 2} for a technical description of the Adjei Method).
These methods appeared to provide an estimate the parasite intensity at which
50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
population that was lost (SI 2). However, to implement this method the observed data must be
modified to fit the GLM framework and subjectively binned when mean
infection intensity is high or sample sizes are small.

After 30 years, and despite clear limitations \citep{McCallum2000a}, these
methods (particularly the Crofton Method) are still discussed among
parasitologists and are the primary techniques for examining population level
impacts of parasitism using parasite intensity data. In these methods, PIHM can
only be identified by visually examining plots and, with no clear decision
rule; it can be difficult to determine the significance of PIHM across
different host-parasite systems. The survival function produced by the Adjei
Method offers one solution; however, this method requires manipulating the
original data and has never been thoroughly tested.

Estimating parasite impacts on host population using intensity data should only
be done if methods exist to permit unbiased and accurate estimates of PIHM. In
this study, we first propose a novel method for calculating PIHM. We
next use simulations to compare our method with the previous Adjei
Method to test the ability of both methods to (1) detect occurrence of PIHM and (2) estimate the
lethal parasite load ($LD_{50}$) and the associated survival function.  We then
apply both methods to real datasets previously used in PIHM analyses and
compare results. Finally, we discuss the limitations of inferring PIHM
from distributional data and whether any method for inferring PIHM has a place
in quantitative parasitology.

@@

Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
require binning or data alteration, reduces the number of
parameters to be estimated, and allows the significance of PIHM to be determined
using standard statistical techniques.

As with all previously proposed methods for estimating PIHM, the
Likelihood Method first assumes that the pre-mortality distribution of parasites across hosts follows a
negative binomial distribution $g(x; \mu_p, k_p)$, where $\mu_p$ is the mean parasite intensity in hosts before mortality and $k_p$ is the parasite aggregation before mortality (smaller $k_p$ leads to more aggregation).   Previous methods have also required a parameter $N_p$ specifying the total number of hosts before mortality \citep{Crofton1971a,Adjei1986}, but this is not a necessary parameter in the Likelihood Method.

The second assumption of the Likelihood Method is that the host survival function, the function specifying the probability of a host surviving with $x$ parasites, takes the form of a logistic curve given by
\begin{equation}
    h(x ; a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
For this equation, the parasite intensity at which 50\% of host experience PIHM ($LD_{50}$) can be calculated by $\exp{a / b}$. With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$.

To estimate these parameters, we first define a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.  Using standard rules of conditional probability this distribution can be written as

@@

Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset. Alternatively, one could apply the Crofton Method to estimate $\mu_p$ and $k_p$ and then find the maximum likelihood estimates of $a$ and $b$ and the corresponding $LD_{50}$.

To estimate the significance of PIHM in a host-parasite system, a
likelihood ratio test can be used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by a negative binomial
distribution.  If PIHM is not significant in the system, the resulting likelihood
ratio statistic should approximately follow a $\chi^2$ distribution with degrees of freedom equal to 2.  We provide the code for implementing this Likelihood Method in SI 3.

@@

To test the ability of the Adjei and the Likelihood Methods to identify whether or not PIHM was occurring in a system, we randomly generated data using the following procedure.  First, we drew $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$.  This represented the dataset observed before mortality. Second, we chose values of $a$ and $b$ and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Third, we drew $N_p$ random numbers from a uniform distribution
between 0 and 1 and if host survival probability was less than this random
number, the host experienced parasite-induced mortality.  The surviving
hosts comprised the dataset that would be obtained in the field, after PIHM.

Using both the pre-mortality and post-mortality simulated datasets,  we assumed
that the values of $N_p$, $\mu_p$, and $k_p$ were known and tested the ability
of both methods to correctly determine whether or not PIHM was occurring.
While this scenario is unrealistic because the parameters $N_p$, $\mu_p$, and
$k_p$ are always unknown, we implemented this scenario as a baseline to
establish the efficacy of the methods independent of the estimates of $N_p$,
$\mu_p$ and $k_p$.  If a method could not correctly predict whether or not PIHM
was occurring under these idealistic conditions, we considered this strong
evidence of the unreliability of this method.  For the Adjei Method, $N_p$,
$\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and
$k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via
the Crofton Method (SI X).

We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:surv}).  For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (Type I error) and correctly identifying PIHM in the post-mortality dataset (power).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500]. $N_p$ is not technically the sample size on which the methods are being
tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\

@@

To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above. For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.  Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.

@@

In the above simulations, we tested the power of the Likelihood Method in the ideal scenario when the pre-mortality parameters were known (or estimated independently via the Crofton Method).  In this simulation, test the ability of the Likelihood Method to correctly identify PIHM and estimate $LD_50$ when the pre-mortality parameters are not known. The simulations in the previous sections showed that the Likelihood method was highly effective at identifying PIHM and the $LD_{50}$ when $\mu_p = 10$ and $k_p = 1$, and as an illustrative example of the limitations of the Likelihood Method, we only use these pre-mortality values when simulating the dataset.

@@

We tested the ability of the Adjei Method and the Likelihood Method to identify
PIHM in 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
given in \cite{Adjei1986} (Table \ref{table:pihm}). In the \cite{Crofton1971a} datasets, the host was
the snail \emph{Gammarus pulex} which acts as the intermediate host for the
acanthocephalan \emph{Polmorphus minutus}. In the \cite{Adjei1986} datasets,
the hosts were two species of lizard fish \emph{Saurida tumbil} and
\emph{Saurida undosquamis} that were infected by the cestode
\emph{Callitetrarhynchus gracilis}.  Males and females of both fish species
were considered separately.

In both studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
Method to estimate the pre-mortality distribution, and then ran the Likelihood
Method and Adjei Method using these pre-mortality parameters.  For the
\cite{Adjei1986} datasets, we followed the same procedure as the authors and
first truncated the data at 2 parasites and then fit the Crofton Method for the
female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
distributions for each species with the results from the females.  Finally, we
applied the Adjei Method and the Likelihood Method to determine whether or not
PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the code provided in SI 3.

@@

The Adjei Method showed highly inflated Type I error rates (i.e. falsely detected
PIHM) for all parameter combinations that we
considered (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  This method also showed the unintuitive pattern of Type I error
rate decreasing as sample size decreased.  This pattern was due to the issue of
binning discussed in the \emph{Introduction} and \emph{SI X}. For small samples sizes, the
applicability of the Adjei Method is compromised without binning the observed
data in some way.  In contrast, the Likelihood Method showed a Type I
error rate at or near the pre-set level of 0.05 for all parameter combinations
and sample sizes considered.

The ability of the Adjei Method to detect PIHM given that it was occurring in a
system (i.e. the power) was close to 1 for larger sample sizes and tended to
decrease as sample size decreased (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  The Likelihood Method had a power close to
one for all parameter combinations and sample sizes considered.  With gradual
survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (solid black lines; Figure \ref{fig:typeI10}, SI2 Figs 1, 2).

% TODO: Need to figure out what to do with the mu = 100 that are being dropped
% because the sample size is too small!

@@

The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
for all combinations of parameters examined in this study (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  Even for
small sample sizes (< 500 hosts), the Likelihood Method's estimate of $LD_{50}$
was largely unbiased, with small biases occurring for host survival functions
that were gradual. The precision of the Likelihood Method's $LD_{50}$ estimates decreased
(increasing coefficient of variation) as sample size decreased for all
parameter combinations we examined (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).

The Adjei Method
always produced biased estimates of the $LD_{50}$ across all parameter combinations (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  For $\mu_p = 10$, the $LD_{50}$
estimates from the Adjei Method were largely unbiased for large samples sizes,
but showed increasing bias as sample size decreased, particularly for steep
survival functions.  As $\mu_p$ increased, the Adjei method
produced biased estimates of $LD_{50}$ across all sample sizes, with bias
increasing as sample size decreased (SI2 Fig 3, 4). The $LD_{50}$ estimates from the Adjei
Method showed large decreases in precision occurring
for the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).

% Set sample size to be small. (50 or 100) and then vary k and mu, Make some sort of heat map to show this?

In terms of the host survival function, the Likelihood Method gave
asymptotically unbiased estimates of $a$ as sample size increased for
all parameter combinations considered (Figure \ref{fig:biasa}, SI2 Fig 5, 6).  However, as sample size decreased, the
Likelihood Method tended to produce severely biased estimates of $a$.
This was generally more pronounced for steeper survival functions and more
aggregated pre-mortality distributions (e.g. Figure \ref{fig:biasa}).  The Adjei Method produced
biased estimates of $a$ across all sample sizes, with the bias
consistently being larger when the survival function was steeper. The bias of
the Adjei Method's estimate of $a$ also increased as $\mu_p$ increased (Figure \ref{fig:biasa}, SI2 Fig 5, 6).

@@

Of the 10 datasets we considered, the previous authors visually detected PIHM
in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
from the pre-mortality parameters of the Crofton Method detected significant
PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
dataset in which the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).

The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate.

@@

Quantifying the impact of parasitism on wild host populations is critical in
both disease modeling and wildlife management. Ideally the relationship between
infection intensity and host survival would be measured experimentally, but for
logistical and ethical reasons, this is often impossible \citep{McCallum2000a}.
Looking for evidence of mortality in parasite distribution data requires the
least amount of information, but is notoriously difficult to implement. The
methodological flaws in the Adjei Method limit its utility, so here we propose
an alternative, likelihood-based, method to estimate host survival and the
$LD_{50}$ from observed parasite intensity data.  At a theoretical level this
method is a significant improvement over the previous methods because it requires fewer parameters,
provides a statistical decision rule for identifying PIHM and does not require
any data manipulation.

@@

We thank X for their useful comments. This work was supported by X.

@@

@@

    \caption{Bias (\textbf{A}) and precision (\textbf{B}) (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes. The pre-mortality parameters for this simulation were $\mu_p = 50$ and $k_p = 0.5$.  The figure shows the simulations for three different host survival functions (gradual, moderate, and steep), each with the same $LD_{50}$.  Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in SI 3 Fig 4 - 9.}

@@

>>>>

\title{Methods for estimating parasite-induced mortality from intensity data and their limitations}
\author{Mark Wilber, Sara Weinstein, and Cherie Briggs}

@@

Infectious agents can have major impacts on animal populations through changing
population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies (Coyne et al. 1989) [get citation], bovine TB \citep{Cox2005}, and
rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
disease, and does not account for the number of infectious agents present.
This method is sufficient for many bacterial and viral agents that reproduce within a
host, however for macroparasites pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013},
wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
wild host populations must succumb to parasitic infections, it is notoriously
difficult to actually quantify parasite-induced host mortality (PIHM) in wild
animal populations \citep{McCallum2000a}.

Ideally, parasite-induced host mortality is
quantified by experimentally infecting and tracking individual hosts in the
wild population; however, for logistical and ethical reasons this method is
rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
relationship between infection intensity and host mortality \citep{Adjei1986}.

\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
observed parasite distribution in the host population to the distribution
predicted in the absence of parasite-induced host mortality. We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1. This method
assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts. Assuming mortality occurs only in these heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
fitting a negative binomial distribution to hosts with lower and lower parasite
loads, and comparing the tail ends of these predicted distributions to the
observed parasite data.

The Crofton Method may be able to detect the presence of PIHM however, but it does not quantify the relationship between infection intensity and host survival
probability. \cite{Adjei1986} suggested that this relationship
could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using those distribution to calculate the
probability of host survival with increasing parasite intensity. To do this,
\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
These methods appeared to provide an estimate for the parasite intensity at which
50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
population that was lost (\emph{SI} 2). However, to implement this method the observed data must be
modified to fit the GLM framework and subjectively binned when mean
infection intensity is high or sample sizes are small.

After 30 years, and despite clear limitations \citep{McCallum2000a}, these
methods (particularly the Crofton Method) are still discussed among
parasitologists and are the primary techniques for examining population level
impacts of parasitism using parasite intensity data. In these methods, PIHM can
only be identified by visually examining plots and, with no clear decision
rule, it can be difficult to determine the significance of PIHM across
different host-parasite systems. The survival function produced by the Adjei
Method offers one solution; however, this method requires manipulating the
original data and has never been tested.

Intensity data should only
be used to estimate parasite impacts on host populations if unbiased and accurate methods exist. In
this study, we first propose a novel method for calculating PIHM. We
next use simulations to compare our method with the previous Adjei
Method to test the ability of both methods to (1) detect occurrence of PIHM and (2) estimate the
lethal parasite load ($LD_{50}$) and the associated survival function.  We then
apply both methods to real datasets previously used in PIHM analyses and
compare the results. Finally, we discuss the limitations of inferring PIHM
from intensity data and whether any method for inferring PIHM has a place
in quantitative parasitology.

@@

Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
require binning or data alteration, reduces the number of
parameters to be estimated, and  uses standard statistical techniques to determine PIHM significance. We provide Python code for implementing the Likelihood Method in \emph{SI} 4.

As with all previously proposed methods for estimating PIHM, the
Likelihood Method first assumes that prior to mortality the parasite distribution is described by a negative binomial $g(x; \mu_p, k_p)$, where $\mu_p$ and $k_p$ are the mean parasite intensity and aggregation before mortality, respectively (smaller $k_p$ indicates more aggregation).   Previous methods required calculating the total number of hosts before mortality ($N_p$) \citep{Crofton1971a,Adjei1986}, however this parameter is not needed in the Likelihood Method.

The Likelihood Method then assumes that the host survival function, which specifies the probability of a host surviving with $x$ parasites, follows the logistic curve given by

\begin{equation}
    h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$ by  first defining a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.  Using standard rules of conditional probability this distribution can be written as

@@

Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset. The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.

To estimate the significance of PIHM in a host-parasite system, a
likelihood ratio test is used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by a negative binomial
distribution.  If PIHM is not significant in the system, the resulting likelihood
ratio statistic should approximately follow a $\chi^2$ distribution with two degrees of freedom.

@@

We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters. First, we created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$. Second, we chose values of $a$ and $b$ and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
between 0 and 1 and if the calculated host survival probability was less than this random
number, the host experienced parasite-induced mortality. The parasite distribution in these simulated surviving hosts is equivalent to the observed parasite distribution in a wild host population that has undergone parasite-induced host mortality.

We used these simulated pre-mortality and post-mortality datasets to test the ability
of both methods to correctly determine whether or not PIHM was occurring when the parameters $N_p$, $\mu_p$ and $k_p$ were known.
  For the Adjei Method, $N_p$,
$\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and
$k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via the Crofton Method.  Although the parameters $N_p$, $\mu_p$, and $k_p$ are always unknown in real systems, a method that fails under these ideal simulation conditions will certainly also fail using less ideal, empirical data.

We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).  For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500]. $N_p$ is not technically the sample size on which the methods are being
tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\

@@

To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above. For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over pre-mortality host population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.  Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.

@@

In the final simulation, we test the ability of the Likelihood Method to correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality parameters are unknown. The previous simulations showed that the Likelihood Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with values of 10 and 1, respectively.  In the best-case scenario where a host-parasite system has these these parameters, we test the power of the Likelihood Method to identify PIHM for gradual, moderate and steep survival functions when the pre-mortality parameters also needed to be estimated.  We perform 500 simulations over a range of different samples sizes following the simulation procedure described above.

@@

We tested the ability of the Adjei Method and the Likelihood Method to identify
PIHM in 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
given in \cite{Adjei1986} (Table \ref{table:pihm}). \citeauthor{Crofton1971a} analyzed infection patterns in the snail \emph{Gammarus pulex} infected with the
acanthocephalan \emph{Polmorphus minutus}. \citeauthor{Adjei1986} analyzed males and females of two species of lizard fish \emph{Saurida tumbil} and
\emph{S. undosquamis} that were infected by the cestode
\emph{Callitetrarhynchus gracilis}.

In both earlier studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
Method to estimate the pre-mortality distribution, and then ran the Likelihood
Method and Adjei Method using these pre-mortality parameters.  For the
\cite{Adjei1986} datasets, we followed the same procedure as the authors and
first truncated the data at 2 parasites and then fit the Crofton Method for the
female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
distributions for each species with the results from the females.  Finally, we
applied the Adjei Method and the Likelihood Method to determine whether or not
PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the code provided in \emph{SI} 4.

@@

The power of the Adjei Method to detect PIHM in a
system was close to unity for larger sample sizes and tended to
decrease as sample size decreased (Figure \ref{fig:question1}B; \emph{SI} 3 Figs 1-3).  The Likelihood Method had a power close to
unity for all parameter combinations and sample sizes considered.  With gradual
survival functions, the power of the Likelihood Method decreased slightly for small samples sizes (Fig. \ref{fig:question1}B, \emph{SI} 3 Figs 1-3).

The Adjei Method showed highly inflated Type I error rates (i.e. falsely detected
PIHM) for all parameter combinations that we
considered (Fig. \ref{fig:question1}A; \emph{SI} 3 Figs 1-3).  This method also showed the unintuitive pattern of Type I error
rate decreasing as sample size decreased.  This pattern was due to the issue of
binning discussed in the \emph{Introduction} and \emph{SI} 2. For small samples sizes, the
applicability of the Adjei Method is compromised without binning the observed
data in some way.  In contrast, the Likelihood Method showed a Type I
error rate at or near the pre-set level of 0.05 for all parameter combinations
and sample sizes considered (Fig. \ref{fig:question1}A; \emph{SI} 3 Figs 1-3).

@@

The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
for all combinations of parameters examined in this study (Fig. \ref{fig:question2}, \emph{SI} 3 Figs 4-6).  Even for
the smallest sample sizes we considered, the Likelihood Method's estimate of $LD_{50}$
was largely unbiased, with small biases occurring for host survival functions
that were gradual. The precision of the Likelihood Method's $LD_{50}$ estimates decreased
(increasing coefficient of variation) as sample size decreased for all
parameter combinations we examined (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6).

The Adjei Method produced biased estimates of the $LD_{50}$ across nearly all parameter combinations (Fig \ref{fig:question2}, \emph{SI} 3 Figs 4-6)).  For $\mu_p = 10$, the $LD_{50}$
estimates from the Adjei Method were largely unbiased for large samples sizes, but as $\mu_p$ increased, the Adjei Method
produced biased estimates of $LD_{50}$ across all sample sizes, with bias
increasing as sample size decreased (Figure \ref{fig:question2}, SI2 Fig 4-6). The $LD_{50}$ estimates from the Adjei
Method showed large decreases in precision with the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:question2}, SI2 Fig 4-6).

In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameters when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig. \ref{fig:question2}, \emph{SI} Fig 7 - 9) The Adjei Method produced
biased estimates of the host survival function across all sample sizes, with the bias
consistently being larger when the survival function was steeper and $\mu_p$ was larger (Fig \ref{fig:question2}, \emph{SI} 3 Figs 7-9).

\subsection*{Detecting PIHM with unknown pre-mortality parameters}

When all pre-mortality parameters were jointly estimated, the Likelihood Method had a power of greater than 0.8 when the survival function was moderate and steep for host sample sizes of 424 and 83 respectively (Figure \ref{fig:real_power}).  When the host survival function was gradual, the Likelihood Method never had a power greater than 0.8 for any post-mortality samples sizes we considered.

@@

Of the 10 datasets we considered, the previous authors visually detected PIHM
in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
from the pre-mortality parameters of the Crofton Method detected significant
PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
dataset in which the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$). The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results that the Adjei Method has a high Type I error rate.

@@

Quantifying the impact of parasitism on wild host populations is critical for managing wildlife populations and understanding parasite transmission. Ideally the relationship between
infection intensity and host survival would be measured experimentally, but for
logistical and ethical reasons, this is often impossible \citep{McCallum2000a}.
Looking for evidence of mortality in parasite distribution data requires the
least amount of information, but is notoriously difficult to implement. The
methodological flaws in the Adjei Method limit its utility, so here we propose
an alternative, likelihood-based, method to estimate host survival and the
$LD_{50}$ from observed parasite intensity data.  This
method is a significant improvement over the previous methods because it requires fewer parameters,
provides a statistical decision rule for identifying PIHM and does not require
any data manipulation.

@@

@@

\label{fig:question1}

@@

    \caption{Bias and precision (coefficient of variation) for the Likelihood Method and Adjei Method estimates of the $a$ parameter and the $LD_{50}$ of the host survival function based on simulated PIHM data over a range of post-mortality sample sizes. The pre-mortality parameters for this simulation were $\mu_p = 50$ and $k_p = 0.5$.  The figure shows the simulations for three different host survival functions (gradual, moderate, and steep), each with the same $LD_{50}$.  Bias and precision results of $LD_{50}$ and $a$ for all other parameter combinations can be found in SI 3 Fig 4 - 9.}

@@

\begin{figure}
    \includegraphics[width=\textwidth]{/Users/mqwilber/Repos/parasite_mortality/results/figure3_for_manuscript.pdf}

    \caption{The power of the Likelihood Method to detect PIHM for gradual, moderate, and steep survival functions when all four parameters $\mu_p$, $k_p$, $a$, and $b$ were jointly estimated. The curves were generated from 500 simulations for 10 pre-mortality population sizes, $N_p$.}

    \label{fig:real_power}
